---
title: "day12"
author: "Elisa"
date: "5 March 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(lme4)
require(ggplot2)
require(dplyr)
require(tidyr)
require(DHARMa)
```
Exercises Day 12

1) Load the data "beetles.txt":
 
Background: 
The national forest agency has asked you for advice - research has suggested that bark beetle attacks will be reduced if 
the climate becomes wetter under climate change, as suggested for the region we are looking at here. 
The forest agency measured bark beetle counts and precipitation over 20 years on 50 different plots across an altitudinal gradient. 

Exercise:
Analyse the data and give your conclusion regarding the effects of precipitation. 

```{r}
beetles <- read.delim("~/mixed effect models/day12/beetles_data.txt")

```


### Investigate the data

```{r}
str(beetles)
xtabs(~plot+year, beetles)

```

some plots before starting


```{r}


beetles %>% ggplot(aes(x=precipitation, y=beetles, color=factor(plot))) +
  geom_point()

beetles %>% ggplot(aes(x=precipitation, y=beetles, color=factor(plot))) +
  geom_point() +
  facet_wrap(~year)

beetles %>% ggplot(aes(x=year, y=beetles, group= factor(plot), color=factor(plot))) +
  geom_point()+
  geom_line()

beetles %>% ggplot(aes(x=year, y=beetles, group= factor(plot), color=factor(plot))) +
  geom_point()+
  geom_line()+
  facet_wrap(~plot)

```



Count data- poisson distribution. 

One option is to fit a model where year is a random effect. In this case, we suppose that years are independent, and that the particular characteristichs of that year can be seen as a factor that groups the data
```{r}

model1 <- glmer(beetles~altitude+precipitation+(1|year)+(1|plot), family=poisson, data=beetles)
summary(model1)

model1.b <- glmer(beetles~altitude+precipitation+year+(1|year)+(1|plot), family=poisson, data=beetles)
summary(model1.b)
```

Another option is to think that year might be a continious variable. We could add year as a fixed effect and allow for random slopes, for example. 
When trying to add the random slope the following warning appears

- Can not converge 

or 

- Warning message:
In checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv,  :
Model is nearly unidentifiable: large eigenvalue ratio - Rescale variables?

Then we rescale the predictors, then not predictors with higher numbers are more important in the model


```{r}

model2 <- glmer(beetles~scale(altitude)+scale(precipitation)+scale(year)+(scale(year)|plot), family=poisson, data=beetles)
summary(model2)

```

The random slope does not explain so much variability. 

Actually, for me, looking at the data, it makes more sense to use year as a categorical ranfom effect. There is not a trend allong the years but more a particular situation each year. 
Year is highly significant. Then the desition on weather it is a random or fixed effects depends if it will be included between the fixed effects. 

### Diagnostics of the model

```{r}
ranefects <- ranef(model1) # random effects for the groups 
hist(ranefects$plot[,1])
hist(ranefects$year[,1]) # random effects should be normally distr+ibuted

#qqmath(ranef(model1)) #for normality of random effects # also looks good

```


We can also check the distr+ibution fo the residuals- pearson residual

```{r}
# using Pearson residuals:
rp <- residuals(model1, type = "pearson") # get pearson residuals
link <- predict(model1, type = "link") # prediction on the link scale, quite overdispersed
plot(rp)
plot(link)
plot(rp ~ link)  
```

Using DAHRma package


```{r}


out<- simulateResiduals(model1)
plot(out)
testDispersion(out) # not really overdispersed
testZeroInflation(out) # heavily zero-inflated but not overdispersed


plotResiduals(beetles$altitude, out$scaledResiduals) # altitude makes a problem
plotResiduals(beetles$precipitation, out$scaledResiduals)

testTemporalAutocorrelation(out) # and finally some test for temporal autocorrelation: looks good
```

there is mimsspecification of beetles altitude. LEts try adding quadratic term

```{r}
model2 <- glmer(beetles~poly(altitude, 2)+precipitation+(1|year)+(1|plot), family=poisson, data=beetles)
summary(model2)
```
The square term is significant. So lets check the model again- significant. 

```{r}
# using Pearson residuals:
rp <- residuals(model2, type = "pearson") # get pearson residuals
link <- predict(model2, type = "link") # prediction on the link scale, quite overdispersed
plot(rp)
plot(link)
plot(rp ~ link)
```


```{r}

out<- simulateResiduals(model2)
plot(out)
testDispersion(out) # not really overdispersed
testZeroInflation(out) # heavily zero-inflated but not overdispersed


plotResiduals(beetles$altitude, out$scaledResiduals) # altitude makes a problem
plotResiduals(beetles$precipitation, out$scaledResiduals)

testTemporalAutocorrelation(out)

```
#OH!!!!!!  really much better naw!!!
